Seasonality of submesoscale dynamics in the Kuroshio Extension

### 2016-07-10 writeup/rocha_etal.tex
Small fixes.

\altaffiltext{1}{Scripps Institution of Oceanography, University of California San Diego, La Jolla, CA, USA}
\altaffiltext{2}{Earth Sciences Division, Jet Propulsion Laboratory, California Institute of Technology, Pasadena, CA, USA}

@@

Inertia-gravity waves, on the other hand, peak in late summer/early fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation in the dominant upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late winter/early ,
whereas submesoscale inertia-gravity waves prevail in late summer/early fall.
Because submesoscale turbulence is weakest in late summer/early fall,
the present results indicate that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have presented strong mesoscale and submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N winter \ref{fig1}a). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late spring/early summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the upper-ocean density structure is well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

\section{Statistics of the surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity gradient tensor
\begin{equation}
\left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right]
\end{equation}
using a centered
second-order finite difference scheme. We then diagnose
the vertical vorticity, and horizontal divergence
\begin{equation}
\zeta \equiv v_x - u_y\, ,
\end{equation}
lateral rate of strain
\begin{equation}
  \alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}\,,\\
\end{equation}
and horizontal divergence
\begin{equation}
\delta \equiv u_x + v_y\, .
\end{equation}
These diagnostics highlight the submesoscale structures in the flow
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early spring
(April 15) and fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained;
its local maximum and RMS are both smaller than $0.5f$.
Indeed, the vorticity and rate of strain are strong in winter (Figure \ref{fig2}):
in both simulations, the RMS vorticity and strain are about twice as large in
late winter/early spring
than in late summer/early fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ increases the wintertime RMS vorticity
and strain by about 40$\%$.

@@

The bulk of vertical vorticity and strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses superinertial
motions and reduces the RMS
vorticity by less than 40$\%$ and the RMS strain by less than 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.22 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late summer/early fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as in summer. Submesoscale fronts and eddies evolve
relatively fast, and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
Figure \ref{fig2}c) also shows that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (Figures \ref{fig3}a-b). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, although
 the vorticity skewness reduces from 1.4 to 1.13 from hourly to daily-averaged.
 The April results are characteristic of winter, indicating that wintertime submesoscale
 surface velocity is strongly dominated by submesoscale turbulence --- this is true even for
 superinertial submesoscale currents. The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
winter (Figure \ref{fig3}e). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; Figure \ref{fig3}f).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in summer) and one with a very steep slope.
That the submesoscale dynamics in October are mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly fields
winter (Figure \ref{fig3}g), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

 Time series of PDFs of vorticity and divergence (supplemental material) show
 a strong oscillation between these two regimes. In late winter/early spring
 the vorticity is strongly positively skewed, whereas the divergence is moderately
 negatively skewed as predicted by frontogenesis \citep{capet_etal2008a,mcwilliams2016}
. In late summer/early fall, the divergence is stronger, but PDFs are much less skewed,
consistent with linear inertia-gravity waves.

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly (the solid curves)
velocity snapshots are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016} who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

The inertia-gravity waves project on the sea-surface.
There is a dramatic difference between the spectra
based on hourly and daily-averaged SSH in October (Figure \ref{fig4}b): at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by super-inertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface inertia-gravity waves conspire to yield weak seasonality
in the spectra of KE and SSH variance based on hourly fields.

\section{Summary and Conclusion}
Our work adds to recent studies that presented modeling \citep{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. In particular, our main finding is that in global simulations with embedded tides the
near-surface inertia-gravity waves in the Kuroshio Extension undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean.

\cite{dasaro1978} showed that velocity of linear internal waves
in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
changes seasonally. In particular the baroclinic modes are
significantly more surface intensified in late summer/early fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore one expects a strong seasonality in the near-surface
expression of internal tides and
other small-scale internal waves generated through internal tide interactions.
Near-inertial wave generation peaks in winter \citep[e.g.,][]{alford2003}, but
those waves project on large horizontal scales \citep[e.g, ][]{qi_etal1995}.

>>>>

\altaffiltext{1}{Scripps Institution of Oceanography, University of California San Diego, La Jolla, CA, USA.}
\altaffiltext{2}{Earth Sciences Division, Jet Propulsion Laboratory, California Institute of Technology, Pasadena, CA, USA.}

@@

Inertia-gravity waves, on the other hand, peak in late summer/early fall,
when the upper ocean is strongly stratified. Thus there exists a
strong seasonal modulation in the dominant upper-ocean submesoscale dynamics:
submesoscale turbulence dominates the upper-ocean dynamics in late winter/early,
whereas submesoscale inertia-gravity waves prevail in late summer/early fall.
Because submesoscale turbulence is weakest in late summer/early fall,
the present results indicate that inertia-gravity waves account for most of the
summertime submesoscale SSH variability.

@@

To study seasonal variations in the upper-ocean dynamics, we focus on the northwest
Pacific, in the vicinity of the Kuroshio
Extension, where previous studies have presented strong mesoscale and submesoscale seasonality
\citep{sasaki_etal2014,qiu_etal2014}.
We analyze a sub-domain of the LLC4320 and LLC2160 simulations of about 2000 km$^2$
spanning 155-175$^\circ$E; 25-40$^\circ$N (Figure \ref{fig1}). The stratification
in this mesoscale-rich subtropical region undergoes a vigorous seasonal cycle: wintertime-enhanced
small-scale turbulence de-stratifies the upper ocean, yielding mixed layers
as deep as 300m. In late spring/early summer enhanced solar radiation and mixed-layer
instabilities re-stratify the upper ocean, yielding mixed layers as shallow as 40m.
Fundamentally, the upper-ocean density structure is well-captured by both LLC simulations:
a comparison with Argo climatology shows that both simulations skillfully represent the Kuroshio
Extension stratification and its seasonal variability  (supplemental material).

@@

\section{Statistics of the surface lateral velocity tensor}
To study the seasonality in the surface velocity, we calculate the lateral velocity gradient tensor
\begin{equation}
\left[\begin{matrix} u_x & u_y\\ v_x&v_y \end{matrix}\right]
\end{equation}
using a centered
second-order finite difference scheme. We then diagnose
the vertical vorticity
\begin{equation}
\zeta \equiv v_x - u_y\, ,
\end{equation}
lateral rate of strain
\begin{equation}
  \alpha \equiv [(u_x-v_y)^2 + (u_y + v_x)^2]^{1/2}\,,\\
\end{equation}
and horizontal divergence
\begin{equation}
\delta \equiv u_x + v_y\, .
\end{equation}
These diagnostics highlight the submesoscale structures in the flow
\citep[e.g.,][]{capet_etal2008a,shcherbina_etal2013}.

Figures \ref{fig1}b-c show snapshots of vertical vorticity $\zeta$ in early spring
(April 15) and fall (October 15) in the LLC4320 (1/48$^\circ$) simulation.
The model solutions depict seasonality in vorticity: large values of
fine-grained vertical vorticity are observed in early spring with maximum
values as large as $4f$, where $f$ is
the local planetary vorticity, and root-mean-square (RMS) of about $0.4f$. In early
fall, the situation is the opposite: the vertical vorticity is relatively coarse-grained;
its local maximum and RMS are both smaller than $0.5f$.
Indeed, the vorticity and rate of strain are strongest in wintertime (Figure \ref{fig2}):
in both simulations, the RMS vorticity and strain rate are about twice as large in
late winter/early spring
than in late summer/early fall. Because the wintertime vorticity and strain rate
are dominated by the smallest scales in the flow (the KE spectra is shallower than
a $-3$ power-law in winter), increasing the resolution from
1/24$^\circ$ to 1/48$^\circ$ increases the wintertime RMS vorticity
and strain by about 40$\%$.

@@

The bulk of vertical vorticity and strain rate are associated with subinertial flows
($T_f = 2\pi/f_0\approx 23.5$h, where $f_0$ is the inertial frequency at
the mean latitude): daily-averaging the velocity fields suppresses super-inertial
motions and reduces the RMS
vorticity by 40$\%$ and the RMS strain by 10$\%$; the seasonal
cycle remains strong (see
dashed lines in figures \ref{fig2}a-b). Most of this seasonal cycle is associated
with submesoscale flows: smoothing the velocity fields
with a Hanning filter with cut-off scale of 100 km dramatically reduces the RMS
vorticity and strain rate. The reduction in variance is about 80$\%$ in winter,
yielding RMS vorticity and strain  rate roughly consistent with the diagnostics from
AVISO gridded geostrophic velocities (compare red lines to black lines in figures \ref{fig2}a-b).
The picture that emerges is consistent with recent studies: shallow baroclinic
instabilities energize the submesoscales in late winter, drawing from the available
potential energy stored in deep mixed layers \citep{sasaki_etal2014,callies_etal2015,callies_etal2016}.

The seasonal cycle of the horizontal divergence, however, showcases the complexity
of the upper-ocean annual variability. If submesoscale eddies and fronts dominated
the near-surface variability all year, then the seasonal cycle of horizontal divergence,
vertical vorticity, and lateral strain rate would be in phase \citep[e.g.,][]{sasaki_etal2014}.
While there is a clear wintertime peak in divergence of daily-averaged velocity
(see dashed lines in figure \ref{fig2}c; RMS divergence $\sim0.22 f$ in the 1/48$^\circ$
simulation), the hourly fields show
a stronger enhancement of
lateral divergence in late summer/early fall (RMS divergence $\sim0.22 f$ the 1/48$^\circ$
simulation). Because the 1/48$^\circ$ simulation better resolves
smaller-scale submesoscale flows, a secondary RMS divergence peak in winter  is
nearly as strong as in summer. Submesoscale fronts and eddies evolve
relatively fast, and there is no clear
temporal and spatial scale separation between those motions and inertia-gravity waves
\citep{mcwilliams2016}:
daily-averaging the velocity fields efficiently suppresses the summertime horizontally
divergent flows, but also reduces the wintertime lateral divergence by about 50$\%$.
Figure \ref{fig2}c also shows that most of the lateral divergence is associated
with submesoscale flows: smoothing the velocity fields with a 100km-cut-off
suppresses more than 80$\%$ of the RMS divergence.

@@

The April vorticity-strain jPDF has a shape characteristic of submesoscale turbulence
 (Figures \ref{fig3}a-b). The alignment of vorticity and strain $\alpha \sim \pm\zeta$
 with strong positive skewness are fingerprints of submesoscale fronts
 \citep{shcherbina_etal2013,mcwilliams2016}. The shape of the vorticity-strain
 jPDF is similar for hourly, daily-averaged, although
 the vorticity skewness reduces from 1.4 to 1.13 from hourly to daily-averaged.
 The April results are characteristic of winter, indicating that wintertime submesoscale
 surface velocity is strongly dominated by submesoscale turbulence.
 The hourly velocity fields are significantly
ageostrophic as depicted by the jPDF of vorticity-Laplacian of SSH
winter (Figure \ref{fig3}e). Even in April, only the daily-averaged fields
are largely
in  geostrophic balance (the jPDF of daily-averaged vorticity vs. Laplacian of SSH
is an ellipse with large eccentricity and main axis tilted by 45$^\circ$; Figure \ref{fig3}f).

The October vorticity-strain jPDF shows much weaker skewness (the vorticity skewness
is 0.68 and 0.67 for hourly and daily-averaged velocities). The shape of the
vorticity-strain jPDF  appears to be a combination of two half-ellipses centered
about $\zeta=0$, one with a 45$^\circ$ slope (characteristic of submesocale fronts
that persist in summer) and one with a very steep slope.
That the submesoscale dynamics in October are mainly unbalanced is clearly depicted
in the shape of jPDF of vorticity-Laplacian of SSH for hourly fields
winter (Figure \ref{fig3}g), which is an ellipse aligned in the vertical axis.
Of course daily-averaging the model suppresses the ageostrophic, super-inertial flows, and
the daily-averaged flow is essentially geostrophic as depicted by the 45$^\circ$-tilted
ellipse in the vorticity-Laplacian of SSH jPDF.

 Time series of PDFs of vorticity and divergence (supplemental material) show
 a strong oscillation between these two regimes. In late winter/early spring
 the vorticity is strongly positively skewed, whereas the divergence is moderately
 negatively skewed as predicted by frontogenesis \citep[e.g., ][]{capet_etal2008a,mcwilliams2016}
. In late summer/early fall, the divergence is stronger, but PDFs are much less skewed,
consistent with linear inertia-gravity waves.

@@

Figure \ref{fig4}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the April and October spectra based on hourly (the solid curves)
velocity snapshots are nearly indistinguishable from each other within 95$\%$ confidence level.
Consistent with the results of
\citet{rocha_etal2016}, who analyzed a 5-month output of the LLC4320 simulation in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales smaller than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig4}a). But this suppression is dramatic in October, when the inertia-gravity waves peak. At scales
smaller than 100 km, 39$\%$ of the surface KE in April is accounted for by super-inertial
flows as opposed to 79$\%$ in October. The seasonality of subinertial submesoscale flows
is more dramatic than the seasonal cycle of the total flow. This is consistent
with the results of \citet{sasaki_etal2014}, which are
based on daily-averaged velocity fields of a different model without tidal forcing
(P. Klein, personal communication).

The inertia-gravity waves project on the sea-surface.
There is a dramatic difference between the spectra
based on hourly and daily-averaged SSH in October (Figure \ref{fig4}b): at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At these scales, 33$\%$ of the SSH variance in April is accounted for by super-inertial
flows as opposed to 83$\%$ in October! Curiously, the out-of-phase seasonal cycle of submesoscale
turbulence and near-surface submesoscale inertia-gravity waves conspire to yield weak seasonality
in the spectra of KE and SSH variance based on hourly fields.

\section{Summary and Conclusion}
Our work adds to recent studies that presented modeling \citep{sasaki_etal2014}
and observational \citep{callies_etal2015,buckingham_etal2016} evidence of vigorous seasonality in
submesoscale turbulence. In particular, our main finding is that in global simulations with embedded tides the
near-surface submesoscale inertia-gravity waves in the Kuroshio Extension undergo
a strong seasonal cycle that is out-of-phase with the seasonal cycle of
submesoscale turbulence. We conjecture that the summertime dominance of inertia-gravity waves
 \citep{callies_etal2015} is a consequence both of suppression of submesoscale turbulence and
enhancement of inertia-gravity waves due to re-stratification of
the upper ocean.

\cite{dasaro1978} showed that the velocity of linear internal waves
in the mixed-layer strongly depends on the density jump at the mixed-layer
base, with largest mixed layer velocities when the jump is strongest.
In summer, the shallow mixed layer overlays a strong seasonal pycnocline,
and thus the internal waves projection onto the mixed-layer may be stronger
according to \cite{dasaro1978}'s arguments. An
alternative explanation is that the shape of the baroclinic modes
changes seasonally. In particular the baroclinic modes are
significantly more surface intensified in late summer/early fall
 (supplemental material).
If the internal wave source has weak seasonal dependence, then a strong
seasonal cycle in the near-surface expression of internal waves is expected.
Internal tide generation does not show a seasonal modulation \citep[e.g.,][]{alford2003},
and therefore one expects a strong seasonality in the near-surface
expression of internal tides and
other submesoscale inertia-gravity waves generated through internal tide interactions.
Near-inertial wave generation peaks in winter \citep[e.g.,][]{alford2003}, but
those waves project on large horizontal scales \citep[e.g, ][]{qi_etal1995}.
